---
title: "rexcelbridge: Bridge Excel Add-in Formulas and RTD Feeds into R (Windows)"
author: |
  Francis Tsiboe, Agricultural Risk Policy Center, North Dakota State University, Fargo, ND 58102  

output:
  github_document:
    toc: false
    toc_depth: 3
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  cache = FALSE,
  echo = TRUE,
  fig.path = "man/figures/",
  out.width = "100%"
)

if (!requireNamespace("badger", quietly = TRUE)) install.packages("badger")

library(badger)
library(knitr)
library(devtools)

# useful during dev
if (requireNamespace("devtools", quietly = TRUE)) {
  devtools::load_all(quiet = TRUE)
  suppressWarnings(try(devtools::document(), silent = TRUE))
}
```

<!-- badges: start -->
[![Project Status: Active – Stable and actively developed](https://www.repostatus.org/badges/latest/active.svg)](https://www.repostatus.org/#active)
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![R-CMD-check](https://github.com/ftsiboe/rexcelbridge/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/ftsiboe/rexcelbridge/actions/workflows/R-CMD-check.yaml)
[![Codecov test coverage](https://codecov.io/gh/ftsiboe/rexcelbridge/graph/badge.svg?token=NP14WU2YK4)](https://codecov.io/gh/ftsiboe/rexcelbridge)
![License: GPL v3](https://img.shields.io/badge/License-GPLv3-blue.svg)
![R >= 4.0](https://img.shields.io/badge/R-%3E=4.0-blue)
[![Contributor Covenant](https://img.shields.io/badge/Contributor%20Covenant-2.1-4baaaa.svg)](code_of_conduct.md)
<!-- badges: end -->

**Acknowledgment:** This package grew out of workflow needs to bridge **Excel RTD/data add-ins** (e.g., DTN ProphetX, Bloomberg) with reproducible **R** pipelines on Windows.

**Abstract:** `rexcelbridge` provides **vendor-neutral** helpers to execute Excel formulas from R via **COM automation**. It supports single-cell and spilled-array results and includes a pluggable “readiness” predicate to treat transient add-in states (e.g., *“Wait”*, *“Loading…”*) as not yet ready. This enables scripted retrieval of market/pricing and other plugin-delivered data directly into R, while keeping provider logic inside Excel.

**Keywords:** Excel COM; RTD; Bloomberg; ProphetX; automation; Windows

---

### 🚦 Requirements (read first)

- **Windows + Microsoft Excel** installed
- **Matching bitness**: R and Office must both be 64-bit *or* both 32-bit
- Appropriate **Excel add-ins installed/enabled** (e.g., *Bloomberg*, *ProphetX*) if you use their formulas
- An **interactive desktop session** (RTD often won’t refresh under headless services)

---

## Installation

```{r installation, echo=TRUE, eval=FALSE, message=FALSE, warning=FALSE}
# install from a local folder during development
devtools::install_github("ftsiboe/rexcelbridge",force = TRUE,upgrade = "never")
```

---

## Quick start

Plain Excel (no vendor add-in needed):

```{r plain excel, echo=TRUE, eval=FALSE, message=FALSE, warning=FALSE}
library(rexcelbridge)
rb_eval_single("=SUM(2,3,5)", visible = TRUE)
```

DTN ProphetX (RTD) example for a single cell

```{r prophetX, echo=TRUE, eval=FALSE, message=FALSE, warning=FALSE}

f <- dtn_prophetX_formula(
  symbol = "BEANS.20254", 
  scale = "Weekly", 
  date = "2025-09-08", 
  field = "Open")

rb_eval_single(f, visible = TRUE) # = 9.94

```


Bloomberg Excel add-in examples (requires Bloomberg add-in):

```{r bloomberg, echo=TRUE, eval=FALSE, message=FALSE, warning=FALSE}
# Spot price
rb_eval_single('=BDP("IBM US Equity","PX_LAST")', visible = TRUE)

```

---

## Package workflow

`rexcelbridge` functions act as a small pipeline:

```
rb_start_excel()        # start hidden Excel via COM (Windows)
  └─ rb_ensure_addin()  # (optional) best-effort enable a vendor add-in by name
      └─ rb_eval_single() / rb_eval_block()
            • write formula(s) into a new workbook
            • poll until “ready” (skipping blanks, Excel errors, tokens like "Wait")
            • return a tidy R data.frame
```

You can customize the “ready” detector:

```{r ready-detector, echo=TRUE, eval=FALSE, message=FALSE, warning=FALSE}
pred <- rb_ready_predicate(wait_tokens = c("Wait","Loading...","Please wait"),
                           treat_errors_as_ready = FALSE)

rb_eval_single('=BDP("IBM US Equity","PX_LAST")', ready_fn = pred, visible = TRUE)
```

---

## API overview

```{r overview, echo=TRUE, eval=FALSE, message=FALSE, warning=FALSE}
rb_start_excel(visible = FALSE)

rb_ensure_addin(xl, pattern)                  # e.g., "Bloomberg", "ProphetX"

rb_excel_quote(x)                             # safely quote Excel string args

rb_ready_predicate(wait_tokens = c("Wait","Loading...","N/A"), treat_errors_as_ready = FALSE)

rb_eval_single(formulas,
               timeout_sec = 120,
               visible = FALSE,
               ready_fn = rb_ready_predicate())

rb_eval_block(formula,
              nrow, ncol,
              top_left = c(1L, 1L),
              timeout_sec = 180,
              visible = FALSE,
              ready_fn = rb_ready_predicate())
```

---

## Troubleshooting

- **Add-in not loading:** Open Excel manually, ensure add-in is enabled (Trust Center), then try again.
- **RTD doesn’t update:** Run in a normal **interactive** Windows session (not as a background service).
- **Bitness mismatch:** Reinstall R or Office so both are 64-bit (recommended) or both 32-bit.
- **Entitlements:** Vendor formulas may return `#N/A` without the proper data permissions.
- **Spill sizing:** For `rb_eval_block()`, set `nrow`/`ncol` to match the actual spill (test in Excel first).
- **Transient “Wait”:** Extend `wait_tokens` in `rb_ready_predicate()` to match provider wording.

---

## 📬 Contact

Questions, suggestions, collaborations: **Francis Tsiboe** — <ftsiboe@hotmail.com>
